---
title: "Live Session 3 — International Business Cycles"
subtitle: "Indonesia and Asia-Pacific comparison"
author: "Caleb Tutty"
format:
  html:
    toc: true
    code-fold: false
    df-print: paged
    html-math-method: katex
execute:
  echo: true
  warning: false
  message: false
---

## Purpose

This notebook answers Question 2(c) from Live Session Problem Set 3 using
quarterly real GDP, private consumption, and gross fixed capital formation. It
treats each business-cycle component as the deviation of the logged real series
from an HP-filtered trend and compares Indonesia with a set of Asia-Pacific
economies plus the United States.

The workflow keeps the longest available quarterly history for each country and
uses an unbalanced panel beginning in 1975. After taking logs and subtracting
each country's trend, differences in currency scale do not affect the cycle
measure.

## Data sources and reproducibility

Every downloaded series comes from one of the following public sources. No API
key is required.

- **Quarterly real GDP for the main country panel:** World Bank [Global
  Economic Monitor (GEM)](https://datacatalog.worldbank.org/infrastructure-data/search/dataset/0037798/global-economic-monitor),
  indicator `NYGDPMKTPSAKN` (real GDP in constant domestic currency, seasonally
  adjusted). The notebook retrieves it through the [World Bank Indicators
  API](https://api.worldbank.org/v2/indicator/NYGDPMKTPSAKN?source=15&format=json).
- **Longer quarterly real GDP histories for Australia, South Korea, and the
  United States:** IMF *International Financial Statistics* (IFS), indicator
  `NGDP_R_SA_XDC`, distributed through the public [DBnomics IMF/IFS
  dataset](https://db.nomics.world/IMF/IFS).
- **Private consumption:** IMF IFS real private-sector final consumption,
  indicators `NCP_R_SA_XDC` and, where the source-seasonally-adjusted series is
  unavailable, `NCP_R_NSA_XDC`, from the same [DBnomics IMF/IFS
  dataset](https://db.nomics.world/IMF/IFS).
- **Investment:** IMF IFS real gross fixed capital formation, indicators
  `NFI_R_SA_XDC` and the non-seasonally-adjusted fallback `NFI_R_NSA_XDC`, from
  the same [DBnomics IMF/IFS dataset](https://db.nomics.world/IMF/IFS).

The exact observations used by a normal render are preserved in the local
[GDP snapshot](data/raw/quarterly_real_gdp.csv) and [consumption/investment
snapshot](data/raw/quarterly_real_components.csv). Their companion metadata
files record when the snapshots were retrieved. Setting `REFRESH_DATA=true`
deliberately replaces these files with the latest source observations.

## 1. Setup

```{r}
#| label: setup
required <- c(
  "tidyverse", "zoo", "mFilter", "moments", "scales", "jsonlite", "httr2"
)
missing <- required[!vapply(required, requireNamespace, logical(1), quietly = TRUE)]
if (length(missing)) {
  stop(
    "Missing project dependencies: ",
    paste(missing, collapse = ", "),
    ". Run renv::restore() before rendering."
  )
}

library(tidyverse)
library(zoo)
library(mFilter)
library(moments)
library(scales)

theme_set(theme_minimal(base_size = 12))
```

## 2. Country set and data source

The preferred measure is **real GDP, quarterly, seasonally adjusted**. The
World Bank *Global Economic Monitor* (GEM) supplies a consistent series for all
countries. For Australia, South Korea, and the United States, longer compatible
IMF *International Financial Statistics* series distributed by DBnomics extend the
history through the requested 1975 boundary. The panel is intentionally
unbalanced: countries enter when their quarterly series begins.

```{r}
#| label: country-map
countries <- tribble(
  ~country,        ~iso2, ~iso3, ~source,          ~series_id,          ~group,
  "Indonesia",     "ID",  "IDN", "World Bank GEM", "NYGDPMKTPSAKN",    "Focus",
  "Malaysia",      "MY",  "MYS", "World Bank GEM", "NYGDPMKTPSAKN",    "ASEAN",
  "Thailand",      "TH",  "THA", "World Bank GEM", "NYGDPMKTPSAKN",    "ASEAN",
  "Philippines",   "PH",  "PHL", "World Bank GEM", "NYGDPMKTPSAKN",    "ASEAN",
  "Singapore",     "SG",  "SGP", "World Bank GEM", "NYGDPMKTPSAKN",    "ASEAN",
  "Vietnam",       "VN",  "VNM", "World Bank GEM", "NYGDPMKTPSAKN",    "ASEAN",
  "Australia",     "AU",  "AUS", "IMF IFS via DBnomics", "Q.AU.NGDP_R_SA_XDC", "Asia-Pacific",
  "New Zealand",   "NZ",  "NZL", "World Bank GEM", "NYGDPMKTPSAKN",    "Asia-Pacific",
  "Japan",         "JP",  "JPN", "World Bank GEM", "NYGDPMKTPSAKN",    "Asia-Pacific",
  "South Korea",   "KR",  "KOR", "IMF IFS via DBnomics", "Q.KR.NGDP_R_SA_XDC", "Asia-Pacific",
  "India",         "IN",  "IND", "World Bank GEM", "NYGDPMKTPSAKN",    "Asia-Pacific",
  "China",         "CN",  "CHN", "World Bank GEM", "NYGDPMKTPSAKN",    "Asia-Pacific",
  "United States", "US",  "USA", "IMF IFS via DBnomics", "Q.US.NGDP_R_SA_XDC", "Comparator"
)

countries
```

### Download helper

The World Bank and DBnomics endpoints are public and require no API key.
Downloads use the R package `httr2`. The first render saves a raw-data snapshot
under `data/raw/`; later renders reuse it so source revisions do not silently
change the results. Set `REFRESH_DATA=true` to download and replace the
snapshot deliberately.

```{r}
#| label: download-data
download_file <- function(url, extension) {
  destination <- tempfile(fileext = extension)

  tryCatch({
    httr2::request(url) %>%
      httr2::req_user_agent("LS3-business-cycles/1.0") %>%
      httr2::req_timeout(seconds = 60) %>%
      httr2::req_retry(max_tries = 3) %>%
      httr2::req_perform(path = destination)

    if (!file.exists(destination) || file.size(destination) == 0) {
      stop("The server returned an empty response.")
    }

    destination
  }, error = function(e) {
    unlink(destination)
    warning("Download failed: ", conditionMessage(e), call. = FALSE)
    NULL
  })
}

read_ifs_gdp <- function(country_map) {
  country_mask <- paste(country_map$iso2, collapse = "+")
  url <- paste0(
    "https://api.db.nomics.world/v22/series/IMF/IFS/Q.",
    country_mask, ".NGDP_R_SA_XDC?observations=1"
  )
  destination <- download_file(url, ".json")
  on.exit(if (!is.null(destination)) unlink(destination), add = TRUE)

  if (is.null(destination)) {
    stop("Could not download quarterly real GDP from IMF IFS via DBnomics.")
  }

  payload <- jsonlite::fromJSON(destination, simplifyVector = FALSE)
  docs <- payload$series$docs
  if (is.null(docs) || !length(docs)) {
    stop("DBnomics returned no IMF IFS GDP series.")
  }

  map_dfr(docs, function(doc) {
    values <- vapply(
      doc$value,
      function(value) if (is.null(value)) NA_real_ else as.numeric(value),
      numeric(1)
    )

    tibble(
      iso2 = doc$dimensions$REF_AREA,
      series_id = doc$series_code,
      date = as.Date(unlist(doc$period_start_day)),
      real_gdp = values
    )
  }) %>%
    filter(!is.na(real_gdp), real_gdp > 0) %>%
    inner_join(
      country_map %>% select(country, iso2, iso3, group, source),
      by = "iso2"
    ) %>%
    select(country, iso2, iso3, group, source, series_id, date, real_gdp)
}

read_gem <- function(country_map) {
  country_codes <- paste(country_map$iso3, collapse = ";")
  end_period <- paste0(format(Sys.Date(), "%Y"), "Q4")
  url <- paste0(
    "https://api.worldbank.org/v2/country/", country_codes,
    "/indicator/NYGDPMKTPSAKN?source=15",
    "&date=1975Q1:", end_period,
    "&format=json&per_page=20000"
  )
  destination <- download_file(url, ".json")
  on.exit(if (!is.null(destination)) unlink(destination), add = TRUE)

  if (is.null(destination)) {
    stop("Could not download World Bank GEM quarterly GDP.")
  }

  payload <- jsonlite::fromJSON(destination, simplifyDataFrame = TRUE)
  if (length(payload) < 2 || !is.data.frame(payload[[2]])) {
    stop("World Bank GEM returned an unexpected response.")
  }

  records <- payload[[2]]
  quarter_number <- as.integer(substr(records$date, 6, 6))
  quarter_month <- c("01", "04", "07", "10")[quarter_number]

  tibble(
    iso3 = records$country$id,
    date = as.Date(paste0(substr(records$date, 1, 4), "-", quarter_month, "-01")),
    real_gdp = as.numeric(records$value)
  ) %>%
    filter(!is.na(real_gdp), real_gdp > 0) %>%
    inner_join(country_map, by = "iso3") %>%
    select(country, iso2, iso3, group, source, series_id, date, real_gdp)
}

raw_dir <- file.path("data", "raw")
raw_data_path <- file.path(raw_dir, "quarterly_real_gdp.csv")
raw_metadata_path <- file.path(raw_dir, "quarterly_real_gdp_metadata.csv")
refresh_data <- tolower(Sys.getenv("REFRESH_DATA", "false")) %in% c("1", "true", "yes")

if (refresh_data || !file.exists(raw_data_path)) {
  ifs_countries <- countries %>% filter(source == "IMF IFS via DBnomics")
  gem_countries <- countries %>% filter(source == "World Bank GEM")
  gdp_raw <- bind_rows(
    read_ifs_gdp(ifs_countries),
    read_gem(gem_countries)
  )

  if (nrow(gdp_raw) == 0) {
    stop("No GDP observations were downloaded; the raw-data snapshot was not updated.")
  }

  absent <- setdiff(countries$country, unique(gdp_raw$country))
  if (length(absent)) {
    stop("No quarterly GDP data returned for: ", paste(absent, collapse = ", "))
  }

  dir.create(raw_dir, recursive = TRUE, showWarnings = FALSE)
  write_csv(gdp_raw, raw_data_path)
  write_csv(
    tibble(
      retrieved_at_utc = format(Sys.time(), tz = "UTC", usetz = TRUE),
      sources = paste(sort(unique(gdp_raw$source)), collapse = "; "),
      country_count = n_distinct(gdp_raw$country),
      observation_count = nrow(gdp_raw)
    ),
    raw_metadata_path
  )
} else {
  gdp_raw <- read_csv(
    raw_data_path,
    col_types = cols(
      country = col_character(),
      iso2 = col_character(),
      iso3 = col_character(),
      group = col_character(),
      source = col_character(),
      series_id = col_character(),
      date = col_date(),
      real_gdp = col_double()
    )
  )
}

coverage <- gdp_raw %>%
  group_by(country, source, series_id) %>%
  summarise(first = min(date), last = max(date), n = n(), .groups = "drop")

coverage
```

### Consumption and investment

My choice of consumption and investment datasets come from IMF *International
Financial Statistics* through the public DBnomics API. Source-seasonally-
adjusted series are preferred; when unavailable, a real non-seasonally-adjusted
series is adjusted with a periodic STL (Seasonal and Trend decomposition using LOESS) decomposition. The IMF doesn't publish this data for all countries (e.g. Vietnam).

```{r}
#| label: download-components
component_definitions <- tribble(
  ~component,    ~sa_indicator,    ~nsa_indicator,
  "Consumption", "NCP_R_SA_XDC",  "NCP_R_NSA_XDC",
  "Investment",  "NFI_R_SA_XDC",  "NFI_R_NSA_XDC"
)

extract_dbnomics_components <- function(country_map) {
  country_mask <- paste(country_map$iso2, collapse = "+")
  indicator_mask <- paste(
    c(component_definitions$sa_indicator, component_definitions$nsa_indicator),
    collapse = "+"
  )
  url <- paste0(
    "https://api.db.nomics.world/v22/series/IMF/IFS/Q.",
    country_mask, ".", indicator_mask, "?observations=1"
  )
  destination <- download_file(url, ".json")
  on.exit(if (!is.null(destination)) unlink(destination), add = TRUE)

  if (is.null(destination)) {
    stop("Could not download IMF IFS components from DBnomics.")
  }

  payload <- jsonlite::fromJSON(destination, simplifyVector = FALSE)
  docs <- payload$series$docs
  if (is.null(docs) || !length(docs)) {
    stop("DBnomics returned no IMF IFS component series.")
  }

  candidates <- map_dfr(docs, function(doc) {
    indicator <- doc$dimensions$INDICATOR
    values <- vapply(
      doc$value,
      function(value) if (is.null(value)) NA_real_ else as.numeric(value),
      numeric(1)
    )

    tibble(
      iso2 = doc$dimensions$REF_AREA,
      component = if_else(str_starts(indicator, "NCP_"), "Consumption", "Investment"),
      series_id = doc$series_code,
      source_adjustment = if_else(str_detect(indicator, "_SA_"), "SA", "NSA"),
      date = as.Date(unlist(doc$period_start_day)),
      value = values
    )
  }) %>%
    filter(!is.na(value), value > 0) %>%
    inner_join(
      country_map %>% select(country, iso2, iso3, group),
      by = "iso2"
    )

  selected_series <- candidates %>%
    distinct(country, component, series_id, source_adjustment) %>%
    mutate(priority = if_else(source_adjustment == "SA", 1L, 2L)) %>%
    group_by(country, component) %>%
    slice_min(priority, n = 1, with_ties = FALSE) %>%
    ungroup() %>%
    select(country, component, series_id)

  candidates %>%
    semi_join(selected_series, by = c("country", "component", "series_id"))
}

adjust_component_seasonality <- function(df) {
  df <- arrange(df, date)
  if (first(df$source_adjustment) == "SA") {
    return(mutate(
      df,
      real_value = value,
      adjustment_method = "Source seasonally adjusted"
    ))
  }

  expected_dates <- seq(min(df$date), max(df$date), by = "3 months")
  if (nrow(df) < 20 || !identical(df$date, expected_dates)) {
    warning("Cannot seasonally adjust incomplete series: ", first(df$series_id))
    return(tibble())
  }

  start_year <- as.integer(format(first(df$date), "%Y"))
  start_quarter <- (as.integer(format(first(df$date), "%m")) - 1L) %/% 3L + 1L
  fit <- stl(
    ts(log(df$value), start = c(start_year, start_quarter), frequency = 4),
    s.window = "periodic",
    robust = TRUE
  )

  mutate(
    df,
    real_value = exp(log(value) - as.numeric(fit$time.series[, "seasonal"])),
    adjustment_method = "Periodic STL adjustment of source NSA series"
  )
}

component_data_path <- file.path(raw_dir, "quarterly_real_components.csv")
component_metadata_path <- file.path(raw_dir, "quarterly_real_components_metadata.csv")

if (refresh_data || !file.exists(component_data_path)) {
  components_raw <- extract_dbnomics_components(countries) %>%
    group_by(country, component) %>%
    group_modify(~ adjust_component_seasonality(.x)) %>%
    ungroup() %>%
    mutate(source = "IMF IFS via DBnomics") %>%
    select(
      country, iso2, iso3, group, component, source, series_id,
      source_adjustment, adjustment_method, date, real_value
    )

  dir.create(raw_dir, recursive = TRUE, showWarnings = FALSE)
  write_csv(components_raw, component_data_path)
  write_csv(
    tibble(
      retrieved_at_utc = format(Sys.time(), tz = "UTC", usetz = TRUE),
      source = "IMF International Financial Statistics via DBnomics",
      country_count = n_distinct(components_raw$country),
      series_count = n_distinct(components_raw$series_id),
      observation_count = nrow(components_raw)
    ),
    component_metadata_path
  )
} else {
  components_raw <- read_csv(
    component_data_path,
    col_types = cols(
      country = col_character(),
      iso2 = col_character(),
      iso3 = col_character(),
      group = col_character(),
      component = col_character(),
      source = col_character(),
      series_id = col_character(),
      source_adjustment = col_character(),
      adjustment_method = col_character(),
      date = col_date(),
      real_value = col_double()
    )
  )
}

component_coverage <- crossing(
  country = countries$country,
  component = component_definitions$component
) %>%
  left_join(
    components_raw %>%
      group_by(country, component, adjustment_method) %>%
      summarise(first = min(date), last = max(date), n = n(), .groups = "drop"),
    by = c("country", "component")
  ) %>%
  arrange(country, component)

component_coverage
```

## 3. Make the sample comparable

The analysis window begins in **1975 Q1**, but each country enters when its
quarterly series becomes available. Summary statistics use every observation
available for that country; correlations use overlapping observations only.

```{r}
#| label: comparable-sample
start_date <- as.Date("1975-01-01")
end_date <- max(gdp_raw$date)

series_data <- bind_rows(
  gdp_raw %>%
    transmute(
      country, iso2, iso3, group, component = "GDP", source, series_id,
      adjustment_method = "Source seasonally adjusted", date,
      value = real_gdp
    ),
  components_raw %>%
    transmute(
      country, iso2, iso3, group, component, source, series_id,
      adjustment_method, date, value = real_value
    )
) %>%
  filter(date >= start_date, date <= end_date) %>%
  arrange(component, country, date)
```

## 4. Construct the business-cycle component

For quarterly data, the conventional Hodrick–Prescott smoothing parameter is
$\lambda=1600$. We work with the log of each real expenditure series:

$$
\operatorname{cycle}_{ijt}=100\left[\log(X_{ijt})-\operatorname{trend}_{ijt}\right],
$$

where $j$ denotes GDP, private consumption, or gross fixed capital formation.
The result is approximately the percentage deviation from the component's
estimated trend.

### 4a. Build the HP filter manually

The HP filter chooses a trend $\tau_t$ that balances fit against smoothness. For
a logged series $y_t=\log(X_t)$, the notation means:

- $X_t$ is the observed real GDP, consumption, or investment level in quarter
  $t$.
- $y_t=\log(X_t)$ is that observed level after taking its natural logarithm.
  It is the input to the filter, not something chosen by the filter.
- $\tau_t$ is the unobserved smooth trend in the same log units as $y_t$. It is
  the quantity chosen by the HP calculation.
- $c_t=y_t-\tau_t$ is the estimated cyclical gap. Multiplying it by 100 gives
  the approximate percentage deviation of the observed series from trend.

The filter chooses the complete sequence $\tau_1,\ldots,\tau_T$ by solving

$$
\min_{\tau_1,\ldots,\tau_T}
\left\{
\sum_{t=1}^{T}(y_t-\tau_t)^2
+\lambda\sum_{t=2}^{T-1}
\left[(\tau_{t+1}-\tau_t)-(\tau_t-\tau_{t-1})\right]^2
\right\}.
$$

The first sum penalises a trend that is far from the data. The expression in
the second sum is the change in the trend's growth rate—its second difference—
so that term penalises a trend that bends sharply.

#### Step 1: take natural logs

Logging makes proportional changes comparable across countries and units. A
log gap of $0.01$ is approximately a one-percent gap, which is why the final
cycle is multiplied by 100.

```{r}
#| label: hp-manual-log
lambda_hp <- 1600

indonesia_example <- series_data %>%
  filter(country == "Indonesia", component == "GDP") %>%
  arrange(date) %>%
  transmute(date, real_level = value, log_value = log(value))

indonesia_example %>% slice_head(n = 8)
```

Note: this is Indonesian Rupiah in 2010 Rupiah values.

#### Step 2: represent changes in trend growth with a matrix

Let $\boldsymbol{y}$ and $\boldsymbol{\tau}$ contain all $T$ observations. A
second-difference matrix $D$ has $1,-2,1$ in each successive row, so
$D\boldsymbol{\tau}$ contains
$\tau_{t-1}-2\tau_t+\tau_{t+1}$. Here is the leading corner of that matrix:

```{r}
#| label: hp-manual-difference-matrix
n_periods <- nrow(indonesia_example)
D2 <- diff(diag(n_periods), differences = 2)
D2[1:4, 1:6]
```

The objective can therefore be written compactly as

$$
(\boldsymbol{y}-\boldsymbol{\tau})'
(\boldsymbol{y}-\boldsymbol{\tau})
+\lambda(D\boldsymbol{\tau})'(D\boldsymbol{\tau}).
$$

#### Step 3: solve the first-order condition

Differentiating with respect to $\boldsymbol{\tau}$ and setting the result to
zero gives

$$
(I_T+\lambda D'D)\boldsymbol{\tau}=\boldsymbol{y},
\qquad
\widehat{\boldsymbol{\tau}}
=(I_T+\lambda D'D)^{-1}\boldsymbol{y}.
$$

In code, `solve(A, y)` solves $A\tau=y$ directly; it is preferable to
explicitly calculating the inverse of $A$.

```{r}
#| label: hp-manual-solve
hp_manual <- function(x, lambda = 1600) {
  x <- as.numeric(x)
  n <- length(x)
  if (n < 3 || any(!is.finite(x))) {
    stop("The manual HP filter needs at least three finite observations.")
  }

  D2 <- diff(diag(n), differences = 2)
  system_matrix <- diag(n) + lambda * crossprod(D2)
  trend <- as.numeric(solve(system_matrix, x))

  tibble(
    hp_trend_manual = trend,
    cycle_manual = 100 * (x - trend)
  )
}

manual_indonesia <- bind_cols(
  indonesia_example,
  hp_manual(indonesia_example$log_value, lambda = lambda_hp)
)

manual_indonesia %>%
  select(date, real_level, log_value, hp_trend_manual, cycle_manual) %>%
  slice_head(n = 8)
```

#### Alternative: minimise the loss function directly

The matrix equation is not the only way to compute the answer. We can code the
two terms of the loss function exactly as written and let a numerical optimiser
search over the $T$ possible trend values. The gradient below distributes the
effect of each second difference back to the three trend observations involved
in it; it does not construct $D$ or solve a matrix equation.

```{r}
#| label: hp-loss-optimisation
hp_loss <- function(tau, y, lambda) {
  fit_loss <- sum((y - tau)^2)
  smoothness_loss <- lambda * sum(diff(tau, differences = 2)^2)
  fit_loss + smoothness_loss
}

hp_loss_gradient <- function(tau, y, lambda) {
  gradient <- 2 * (tau - y)
  curvature <- diff(tau, differences = 2)

  for (i in seq_along(curvature)) {
    gradient[i] <- gradient[i] + 2 * lambda * curvature[i]
    gradient[i + 1] <- gradient[i + 1] - 4 * lambda * curvature[i]
    gradient[i + 2] <- gradient[i + 2] + 2 * lambda * curvature[i]
  }

  gradient
}

hp_optim <- function(x, lambda = 1600) {
  result <- optim(
    par = x,
    fn = hp_loss,
    gr = hp_loss_gradient,
    y = x,
    lambda = lambda,
    method = "BFGS",
    control = list(maxit = 5000, reltol = 1e-12)
  )

  list(
    trend = result$par,
    cycle = 100 * (x - result$par),
    convergence = result$convergence,
    objective = result$value
  )
}

optim_indonesia <- hp_optim(
  indonesia_example$log_value,
  lambda = lambda_hp
)

tibble(
  convergence_code = optim_indonesia$convergence,
  minimised_loss = optim_indonesia$objective,
  maximum_cycle_difference_from_matrix = max(
    abs(optim_indonesia$cycle - manual_indonesia$cycle_manual)
  )
)
```

`method = "BFGS"` asks `optim()` to use the
**Broyden–Fletcher–Goldfarb–Shanno** algorithm. BFGS is a quasi-Newton method:
it uses the loss and its gradient to build an improving approximation to the
inverse Hessian—the local curvature of the loss—rather than calculating the
full Hessian at every iteration. It is appropriate here because the HP loss is
smooth, unconstrained, and convex, so it has one global minimum. A convergence
code of zero indicates that `optim()` satisfied its stopping rule. Numerical
optimisation makes the loss definition especially transparent, although the
direct matrix solution is faster and more accurate for this quadratic problem.

#### Why diagonalisation works

There are two related uses of `diag()` in the manual calculation. First,
`diag(T)` creates the identity matrix; taking its second row differences creates
the $D$ matrix whose rows are $[1,-2,1]$. Second, the curvature-penalty matrix
$K=D'D$ can itself be **diagonalised** because it is real and symmetric:

$$
K=Q\,\operatorname{diag}(\mu_1,\ldots,\mu_T)Q',
$$

where the columns of $Q$ are orthonormal eigenvectors and each $\mu_j\geq0$ is
an eigenvalue measuring the curvature of that mode. Substitution into the HP
solution gives

$$
\widehat{\boldsymbol{\tau}}
=Q\,\operatorname{diag}\left(
\frac{1}{1+\lambda\mu_1},\ldots,
\frac{1}{1+\lambda\mu_T}
\right)Q'\boldsymbol{y}.
$$

This works because multiplication in the eigenvector basis becomes simple
element-by-element scaling. Smooth modes have small $\mu_j$ and receive weights
near one; highly curved modes have large $\mu_j$ and are strongly attenuated.
The two zero-curvature modes correspond to a constant and a straight line, so
the HP penalty does not shrink them. This is also the signal-processing
interpretation of the filter: it preserves slow-moving trend modes and assigns
more rapidly changing modes to the cycle.

```{r}
#| label: hp-diagonalisation
hp_diagonalised <- function(x, lambda = 1600) {
  x <- as.numeric(x)
  n <- length(x)
  D2 <- diff(diag(n), differences = 2)
  eig <- eigen(crossprod(D2), symmetric = TRUE)

  # Tiny negative values can arise from floating-point roundoff even though
  # D'D is positive semidefinite.
  curvature <- pmax(eig$values, 0)
  trend_weights <- 1 / (1 + lambda * curvature)
  mode_coefficients <- as.numeric(crossprod(eig$vectors, x))
  trend <- as.numeric(eig$vectors %*% (trend_weights * mode_coefficients))

  list(
    trend = trend,
    cycle = 100 * (x - trend),
    curvature = curvature,
    trend_weights = trend_weights
  )
}

diagonal_indonesia <- hp_diagonalised(
  indonesia_example$log_value,
  lambda = lambda_hp
)

tibble(
  near_zero_curvature_modes = sum(diagonal_indonesia$curvature < 1e-10),
  smallest_trend_weight = min(diagonal_indonesia$trend_weights),
  maximum_cycle_difference_from_direct_solve = max(
    abs(diagonal_indonesia$cycle - manual_indonesia$cycle_manual)
  )
)
```

#### Why is $\lambda=1600$?

$\lambda$ controls the trade-off rather than describing the data itself:

- At $\lambda=0$, the fitted trend follows every observation and the cycle is
  zero.
- As $\lambda$ increases, changes in trend growth become more costly, the trend
  becomes smoother, and more medium-frequency movement is assigned to the
  cycle.
- As $\lambda\rightarrow\infty$, the fitted trend approaches a straight line.

The value 1600 is the convention proposed by [Hodrick and Prescott
(1997)](https://papers.ssrn.com/sol3/papers.cfm?abstract_id=5052) for quarterly
macroeconomic data; it is not estimated from this sample and is not universally
optimal. Sampling frequency matters because a second difference has a
different scale when observations are closer together. [Ravn and Uhlig
(2002)](https://discovery.ucl.ac.uk/id/eprint/18641/) recommend scaling
$\lambda$ by the fourth power of the frequency ratio. Taking quarterly 1600 as
the benchmark gives $1600/4^4=6.25$ for annual data and
$1600\times3^4=129{,}600$ for monthly data. We retain 1600 here because every
series entering the filter is quarterly, and later treat the choice as a
modelling assumption rather than a fact.

### 4b. Apply the filter to every country and component

The matrix calculation above explains the mechanics. For the full panel we use
`mFilter::hpfilter()` and verify that it reproduces the manual calculation.

```{r}
#| label: hp-filter
hp_one_series <- function(df) {
  df <- arrange(df, date)
  if (nrow(df) < 20) return(tibble())

  x <- log(df$value)
  hp <- mFilter::hpfilter(x, freq = lambda_hp, type = "lambda")

  df %>%
    mutate(
      log_value = x,
      hp_trend = as.numeric(hp$trend),
      cycle = 100 * as.numeric(hp$cycle),
      qoq_growth = 100 * (log_value - lag(log_value)),
      qoq_annualised = 400 * (log_value - lag(log_value)),
      yoy_growth = 100 * (log_value - lag(log_value, 4))
    )
}

cycles <- series_data %>%
  group_by(country, component) %>%
  group_modify(~ hp_one_series(.x)) %>%
  ungroup()

manual_package_check <- cycles %>%
  filter(country == "Indonesia", component == "GDP") %>%
  select(date, package_cycle = cycle) %>%
  inner_join(
    manual_indonesia %>% select(date, cycle_manual),
    by = "date"
  ) %>%
  summarise(
    maximum_absolute_difference_percentage_points =
      max(abs(package_cycle - cycle_manual))
  )

manual_package_check
```

The reported maximum difference should be effectively zero (up to floating-
point rounding), demonstrating that the manual linear-system solution and
`mFilter` implement the same HP calculation.

## 5. Indonesia: inspect the cycle

```{r}
#| label: indonesia-cycle-plot
cycles %>%
  filter(country == "Indonesia") %>%
  ggplot(aes(date, cycle, colour = component)) +
  geom_hline(yintercept = 0, linewidth = 0.4) +
  geom_line(linewidth = 0.8) +
  labs(
    title = "Indonesia: output and expenditure cycles",
    subtitle = "HP-filtered log real series, λ = 1600",
    x = NULL,
    y = "% deviation from trend",
    colour = NULL
  )
```

### Growth-rate sanity check

```{r}
#| label: indonesia-growth
cycles %>%
  filter(country == "Indonesia", component == "GDP") %>%
  ggplot(aes(date, yoy_growth)) +
  geom_hline(yintercept = 0, linewidth = 0.4) +
  geom_line(linewidth = 0.8) +
  labs(
    title = "Indonesia: year-on-year real GDP growth",
    x = NULL,
    y = "%"
  )
```

## 6. Statistical properties

The classic statistics are volatility, persistence, and co-movement. Reporting
them by component tests two standard international-business-cycle patterns:
consumption is often smoother than output, while investment is usually more
volatile. We add skewness and kurtosis because large asymmetric shocks can make
the cycles non-Gaussian.

```{r}
#| label: summary-stats
safe_cor_lag1 <- function(x) {
  x <- x[is.finite(x)]
  if (length(x) < 3) return(NA_real_)
  cor(x[-length(x)], x[-1], use = "complete.obs")
}

stats <- cycles %>%
  group_by(country, iso2, group, component) %>%
  summarise(
    observations = sum(is.finite(cycle)),
    cycle_sd = sd(cycle, na.rm = TRUE),
    cycle_mean_abs = mean(abs(cycle), na.rm = TRUE),
    persistence_ar1 = safe_cor_lag1(cycle),
    skewness = moments::skewness(cycle, na.rm = TRUE),
    kurtosis = moments::kurtosis(cycle, na.rm = TRUE),
    mean_yoy_growth = mean(yoy_growth, na.rm = TRUE),
    sd_yoy_growth = sd(yoy_growth, na.rm = TRUE),
    .groups = "drop"
  )

# Compare each component with GDP over exactly the same country-quarter sample.
# This avoids mistaking different start dates for differences in volatility.
gdp_reference <- cycles %>%
  filter(component == "GDP") %>%
  select(country, date, gdp_cycle = cycle)

component_comparison <- cycles %>%
  inner_join(gdp_reference, by = c("country", "date")) %>%
  group_by(country, component) %>%
  summarise(
    comparison_observations = sum(is.finite(cycle) & is.finite(gdp_cycle)),
    relative_volatility_to_gdp = sd(cycle, na.rm = TRUE) /
      sd(gdp_cycle, na.rm = TRUE),
    corr_with_domestic_gdp = cor(cycle, gdp_cycle, use = "complete.obs"),
    .groups = "drop"
  )

stats <- stats %>%
  left_join(component_comparison, by = c("country", "component"))

stats %>% arrange(country, component)
```

### Correlation with Indonesia's cycle

```{r}
#| label: correlations
safe_cor <- function(x, y) {
  paired <- is.finite(x) & is.finite(y)
  if (sum(paired) < 3) return(NA_real_)
  cor(x[paired], y[paired])
}

indonesia_cycles <- cycles %>%
  filter(country == "Indonesia") %>%
  select(date, component, indonesia_cycle = cycle)

correlations <- cycles %>%
  inner_join(indonesia_cycles, by = c("date", "component")) %>%
  group_by(country, iso2, group, component) %>%
  summarise(
    corr_with_indonesia = safe_cor(cycle, indonesia_cycle),
    correlation_observations = sum(is.finite(cycle) & is.finite(indonesia_cycle)),
    .groups = "drop"
  ) %>%
  arrange(component, desc(corr_with_indonesia))

correlations
```

```{r}
#| label: corr-bar
correlations %>%
  filter(country != "Indonesia") %>%
  ggplot(aes(x = reorder(country, corr_with_indonesia), y = corr_with_indonesia)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ component) +
  labs(
    title = "Business-cycle correlation with Indonesia",
    subtitle = "Contemporaneous correlation of HP-filtered real component cycles",
    x = NULL,
    y = "Correlation"
  )
```

## 7. Rolling synchronisation

A single full-sample correlation can hide structural change. The next calculation uses a 20-quarter (5-year) rolling correlation with Indonesia.

```{r}
#| label: rolling-correlation
roll_cor <- function(x, y, width = 20) {
  zoo::rollapplyr(
    cbind(x, y),
    width = width,
    by.column = FALSE,
    fill = NA_real_,
    FUN = function(z) {
      paired <- complete.cases(z)
      if (sum(paired) < 3) return(NA_real_)
      cor(z[paired, 1], z[paired, 2])
    }
  )
}

rolling <- cycles %>%
  filter(component == "GDP") %>%
  select(date, country, cycle) %>%
  pivot_wider(names_from = country, values_from = cycle) %>%
  arrange(date)

rolling_long <- map_dfr(setdiff(names(rolling), c("date", "Indonesia")), function(ctry) {
  tibble(
    date = rolling$date,
    component = "GDP",
    country = ctry,
    rolling_corr = roll_cor(rolling[[ctry]], rolling$Indonesia, 20)
  )
})

rolling_long %>%
  filter(country %in% c(
    "Malaysia", "Thailand", "Philippines", "Singapore", "Vietnam",
    "Australia", "New Zealand", "China"
  )) %>%
  ggplot(aes(date, rolling_corr, group = country)) +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  geom_line() +
  facet_wrap(~ country, ncol = 2) +
  coord_cartesian(ylim = c(-1, 1)) +
  labs(
    title = "Five-year rolling business-cycle correlation with Indonesia",
    x = NULL,
    y = "Correlation"
  )
```

## 8. COVID robustness check

The pandemic is a genuine macroeconomic shock, but its size can dominate standard deviations and correlations. Report both the full sample and a pre-COVID sample.

```{r}
#| label: robustness
stats_by_period <- cycles %>%
  mutate(period = if_else(date < as.Date("2020-01-01"), "Pre-COVID", "2020 onward")) %>%
  group_by(country, component, period) %>%
  summarise(
    cycle_sd = sd(cycle, na.rm = TRUE),
    mean_yoy_growth = mean(yoy_growth, na.rm = TRUE),
    sd_yoy_growth = sd(yoy_growth, na.rm = TRUE),
    .groups = "drop"
  )

stats_by_period
```

## 9. Export data for a D3 visualisation

```{r}
#| label: export
out_dir <- "d3_data"
dir.create(out_dir, showWarnings = FALSE)

cycles %>%
  select(date, country, iso2, group, component, cycle, yoy_growth, qoq_annualised) %>%
  write_csv(file.path(out_dir, "business_cycle_data.csv"))

stats %>%
  left_join(
    correlations %>% select(country, component, corr_with_indonesia, correlation_observations),
    by = c("country", "component")
  ) %>%
  write_csv(file.path(out_dir, "business_cycle_stats.csv"))

rolling_long %>%
  write_csv(file.path(out_dir, "rolling_correlations.csv"))
```
